First Season MWA EoR Power Spectrum Results at Redshift 7

### 2016-06-29 ms.tex
JDB comments

The 21 cm hyperfine transition from neutral hydrogen residing in the IGM during reionization 
offers another promising method to study the EoR. Not only is the 21cm signal a 
\emph{direct} probe of the IGM, but, due to the narrow width of the transition and the 
relationship between observed frequency and line-of-sight distance, it can be used to map 
the full three dimensional space of the epoch. (For reviews, see \citealt{Furlanetto:2006}, 
\citealt{Morales:2010}, \citealt{Zaroubi:2013}, and \citealt{Pritchard:2012}) The first generation of instruments with primary science goals 
to detect the highly redshift 21cm signal have been built, including the GMRT (Giant Metrewave 
Radio Telescope, \citealt{Paciga:2013}), LOFAR (LOw Frequency Array\footnote{http://
www.lofar.org}, \citealt{vanHaarlem:2013, Yatawatta:2013}), PAPER (Precision Array for Probing the Epoch of 
Reionization\footnote{http://eor.berkeley.edu}, \citealt{Parsons:2010}), and the MWA 
(Murchison Widefield Array\footnote{http://www.mwatelescope.org}, \citealt{Tingay:2013, Bowman:2013}). 
Due to relatively low signal to noise, these instruments aim for a statistical measurement of 
the EoR in the form of a cosmological power spectrum. Meanwhile, the second generation 
21cm EoR machines is on its way with the Hydrogen Epoch of Reionization Array (HERA
\footnote{http://reionization.org}, \citealt{DeBoer:2016}), 
and the low frequency Square Kilometer Array (SKA1-Low \citealt{Mellema:2013}),
which will further 
refine the power spectrum measurement in early build-out stages, but will ultimately be 
capable of imaging the ionized bubbles of reionization in its later stages 
\citep{Beardsley:2015,Malloy:2013}.

@@

While the 21 cm EoR signal is priority, the MWA is a general observatory serving several 
science programs beyond cosmology including Galactic and extragalactic surveys, time 
domain astrophysics, solar monitoring, and ionospheric science. The array was thus 
designed with these programs in consideration and the layout was optimized using a 
pseudo-random antenna placement algorithm \citep{Beardsley:2012} to obtain a well 
behaved point spread function while retaining a dense core for EoR sensitivity 
\citep{Beardsley:2013}.\footnote{A drone flyover view of the array layout is available on YouTube at \url{https://youtu.be/yDWdTUzTUMo}.}
A full description of the science capabilities of the array is found in
\citealt{Bowman:2013}.
The high imaging capability of the MWA enables us to calibrate and subtract foregrounds
based on sky models, leveraging the full capability of the observatory for the
EoR experiment. 
This strategy is in contrast to more targeted EoR experiments like PAPER, 
which utilize a highly redundant layout to enhance sensitivity. Redundant arrays can 
exploit symmetries of the instrument for quick calibration and analysis, but have poor 
point spread functions, making foregrounds more difficult to characterize.

@@

An example two dimensional residual power spectrum formed from the three hour golden 
data set is shown in Figure~\ref{fig:example_ps}. The bulk of the residual power is in the 
so-called foreground ``wedge", indicated with the diagonal black lines where the solid line 
corresponds to sky emission at the horizon, and the dashed lines corresponds to emission 
at the edge of the MWA field of view. Above the wedge we see horizontal lines of 
contamination due to the periodic frequency sampling function. These lines are often 
referred to as the coarse band harmonics. We see vertical streaks at high $k_{||}$ and high 
$k_{\perp}$. These are due to sparse $(u,v)$ sampling beyond the dense core of the MWA 
(starting $\sim70\lambda$), and the foreground power mixing to high spectral 
modes. Between the coarse band lines and to the left of the vertical streaks we see regions 
which contain both positive and negative (indicated by blue on the color bar) pixels. These 
regions are where we have successfully isolated the foregrounds and retained a noise-like 
EoR window in the three hour integration.

@@

The two-dimensional power spectrum is a useful figure of merit (FoM) as we improve and 
refine our analysis pipeline. Foregrounds and systematics often manifest with characteristic 
shapes in this space, enabling us to diagnose problems and quantify improvements 
\citep{Morales:2012}. We use the MWA 3 hour ``golden'' data set from August 23, 2013, to 
repeatedly form power spectra to test and refine our analysis. While an exhaustive catalog 
of these improvements would be unsuited for this article (and likely bore the reader), we 
demonstrate the utility of the 2D power spectrum as a FoM with a few key techniques we 
have employed in our analysis.

@@

J.~L.~B.~Line, et al (2016, in review) performed similar analysis, but with an emphasis on the
positional accuracy of catalog entries. By using simulations of visibilities they were able to 
isolate the effect of using perfect versus offset source positions. They confirmed the
importance of a complete and accurate point source catalog in order to retain a clean EoR
window.

@@

Ultimately we need model visibilities to subtract from our data. However, rather than storing 
a model for all observations (different pointings and phase centers), we find it most simple 
to treat the model as a single HEALPix image which can be used to create model visibilities 
for each independent observation. FHD treats each pixel in the image as a point source at 
the pixel location with total flux density equal to the surface brightness of the diffuse image times the area 
of the pixel and creates model visibilities for each snapshot in the same way that it imports 
a catalog of point sources.

@@

\section{A Deep Integration}\label{sec:deep}
Up to this point we have only discussed testing and analysis on the three hour golden data 
set. We next turn towards a deep integration, incorporating the techniques described above. 
We start with all MWA observations taken on the EoR0 field at 182~MHz between August 
23, 2013 and November 29, 2013. This includes 2,780 snapshots, or about 86.5 hours of 
data. We first make data quality cuts, then form power spectra.

@@

We use the jackknife method to filter out bad data and detect patterns. This involves dividing
the data into many subgroups and forming power spectra. The goal is to find observational
parameters which affect the quality of the data, such as the day or time of night.
A powerful grouping is to divide the data into the observation day 
and ``pointing" (which direction the antennas were pointed while tracking the EoR0 field). 
We show an example of one day's worth of per-pointing power spectra in Figure~
\ref{fig:jackknife}. The pointings are labeled sequentially with $-5$ corresponding to five 
pointing steps prior to zenith transit, $0$ corresponding to zenith, and $+4$ corresponding 
to four steps after zenith. 
This numbering scheme is shown graphically in Figure~\ref{fig:pointing}.
Early in the night ($-5$ through $-3$) the bulk of the Galactic plane 
was still above the horizon, and despite being very far from our primary field of view, highly 
contaminated our observations through the sidelobes of the instrument. We then have 
relatively well-behaved pointings, with the EoR window dominated by noise, until the final 
pointing of the night when we can see evidence of the Galaxy rising again indicated by 
strong lines of power at the edge of the foreground wedge. We saw the same contamination 
on all days of observation and decided to cut all $-5$, $-4$, $-3$, and $+4$ pointings from 
our data set. In principle it may be possible to model the Galactic disk well enough to 
account for its presence, but with a lack of sufficient Galactic model, beam model, and the 
analysis tools to include it, we leave this to future work.

@@

Motivated by the hand grading done with power spectrum jackknives, we developed 
another metric to quickly predict power spectrum quality for each snapshot in our data set. 
We do this by forming a delay spectrum \citep{Parsons:2012b} from the raw, uncalibrated 
visibilities. We then calculate an estimated total EoR window power by integrating the 
power above the horizon line and below the first coarse band line. While we would ideally 
use calibrated visibilities for this metric, we have found that the uncalibrated power is 
strongly correlated to the calibrated power, and the cuts we make below are independent of 
calibration. This is encouraging because future analysis could use this metric before 
processing the snapshots, saving valuable computing resources\footnote{Because
calibration involves generating model visibilities for each snapshot, approximately half
the computational cost of the entire imaging pipeline is required to obtain 
calibration solutions.}.

@@

The 32 hour integrated residual power spectra for our three sub-bands and both 
instrumental polarizations are shown in Figure~\ref{fig:2d_ps_subband}. A number of 
features can be seen in these spectra. In all cases the foreground wedge is prominent, with 
extra power at large scales (low $k_{\perp}$). This is an indication that our diffuse model can 
improve for deeper subtraction.

@@

The lowest region of the EoR window, above the horizon line and around $k_{||} \approx 
0.25$ h Mpc$^{-1}$, contains purely positive bins, indicating non-noise-like power. The 
cause for this seeming leakage is yet unknown, but will ultimately limit our integration in this 
analysis. One suspected origin of this contamination was individual snapshots with high 
contamination that slipped through our cuts. However, dividing the data into random 
subsets and comparing results indicated that there is not a small number of offending 
observations, instead the leakage appears to exist at a low level in all the data.

@@

In all bands and polarizations, we are heavily signal dominated in our most sensitive region
(low $k$). This is not surprising based on the 2D spectra we examined earlier. The gaps in
the data are due to the excised coarse band harmonics. We can see the cable reflection line
between the first two gaps in all polarizations and sub-bands. Between the coarse band
harmonic lines, especially at larger $k$, are bins which approach the noise level and are
consistent with zero (notably the North-South polarization for the low and mid bands). The
bin just above the first coarse band harmonic is also consistent with zero in the low band.
These regions are encouraging because they have the potential to continue integrating down
with more data. Improved suppression of the coarse band harmonic and cable reflection lines
will also improve the limits presented here.

@@

The second limited region in our 1D power spectra is between the coarse band harmonics 
where our measured signal approaches the noise level. Additional data will likely reduce the 
noise, and therefore our upper limit, in this regime. However, until the leakage at low $k$ is 
understood, progress at higher $k$ is susceptible to running into the same systematic. 
While a longer integration may improve the upper limit in the short term, the hope for an 
eventual detection lies in understanding the contaminants at low $k$.

Finally, we place our best upper limits in context with the 21cm EoR field as a whole. A 
direct comparison between measurements from different instruments is difficult due to the 
varying methods, redshifts, and scales probed. However, if we approximate the power 
spectrum in units of mK$^2$ to be roughly flat over the scales probed by the MWA, PAPER, 
and GMRT, we can glean a view of the current state of the field. The results from this 
analysis are a modest improvement over the previous best MWA results \citep{Dillon:2015}, 
which is not surprising given that this integration contains about ten times more data, but 
has a strong signal to noise on the power spectrum. Our results are also consistent with the 
40 hour GMRT limit of $\Delta^2\le6.15\times10^4$~mK$^2$ at $k=0.5\, h$~Mpc$^{-1}$ and $z=8.6$ \citep{Paciga:2013}, though 
probe significantly different periods of the EoR. The frontrunner in terms of an upper limit is 
the PAPER experiment \citep{Ali:2015, Jacobs:2015, Parsons:2014}. The PAPER-64 limit of 
$\Delta^2\le502$~mK$^2$ at $z=8.4$ from a 135 day observing campaign is about one and a half orders of magnitude lower 
than our best limit in power units, though again probing a significantly different redshift. 
While the MWA analysis is not yet at that level of maturity, there is room for rapid 
improvement. The red lines in Figure~\ref{fig:1d_ps} indicate the noise level possible 
to reach with these data if systematics can be overcome.

@@

While the analysis presented here has not reached full potential, it represents the deepest power spectrum 
integration to date produced by an imaging pipeline. Imaging involves many difficulties (e.g. 
efficient gridding and mapmaking, foreground modeling), but if systematics can be 
overcome, it has the potential to quickly compete with other more pointed experiments and 
analysis styles. 
More broadly, imaging analyses will be necessary to perform cross-correlation studies
with complimentary probes such as galaxy surveys or other intensity mapping experiments
 \citep[e.g.][]{DeBoer:2016, Vrbanec:2016, Beardsley:2015, Silva:2015, Dore:2014, Lidz:2009}.
Here we have identified a crucial region of power spectrum space that is 
currently contaminated, with suggestions for improvement in future analysis. With improved 
calibration techniques, primary beam models, and understanding of foregrounds, the MWA 
and other imaging analyses will be able to quickly approach the most competitive upper 
limits in the field.

>>>>

The 21 cm hyperfine transition from neutral hydrogen residing in the IGM during reionization 
offers another promising method to study the EoR. Not only is the 21cm signal a 
\emph{direct} probe of the IGM, but, due to the narrow width of the transition and the 
relationship between observed frequency and line-of-sight distance, it can be used to map 
the full three dimensional space of the epoch. (For reviews, see \citealt{Furlanetto:2006}, 
\citealt{Morales:2010}, \citealt{Zaroubi:2013}, and \citealt{Pritchard:2012}) The first generation of instruments with primary science goals 
to detect the highly redshift 21cm signal have been built, including the GMRT (Giant Metrewave 
Radio Telescope, \citealt{Paciga:2013}), LOFAR (LOw Frequency Array\footnote{http://
www.lofar.org}, \citealt{vanHaarlem:2013, Yatawatta:2013}), PAPER (Precision Array for Probing the Epoch of 
Reionization\footnote{http://eor.berkeley.edu}, \citealt{Parsons:2010}), and the MWA 
(Murchison Widefield Array\footnote{http://www.mwatelescope.org}, \citealt{Tingay:2013, Bowman:2013}). 
Due to relatively low signal to noise, these instruments aim for a statistical measurement of 
the EoR in the form of a cosmological power spectrum. 
Complementary to power spectrum experiments are efforts to detect the sky-average global
21-cm signal from the EoR \citep[e.g.][]{Bowman:2010, Sokolowski:2015, Patra:2013, Voytek:2014}.
Meanwhile, the second generation of
21cm EoR interferometers is on its way with the Hydrogen Epoch of Reionization Array (HERA
\footnote{http://reionization.org}, \citealt{DeBoer:2016}), 
and the low frequency Square Kilometer Array (SKA1-Low \citealt{Mellema:2013}),
which will further 
refine the power spectrum measurement in early build-out stages, but will ultimately be 
capable of imaging the ionized bubbles of reionization in its later stages 
\citep{Beardsley:2015,Malloy:2013}.

@@

While the 21 cm EoR signal is priority, the MWA is a general observatory serving several 
science programs beyond cosmology including Galactic and extragalactic surveys, time 
domain astrophysics, solar monitoring, and ionospheric science. The array was thus 
designed with these programs in consideration and the layout was optimized using a 
pseudo-random antenna placement algorithm \citep{Beardsley:2012} to obtain a well 
behaved point spread function while retaining a dense core for EoR sensitivity 
\citep{Beardsley:2013, Bowman:2006}.\footnote{A drone flyover view of the array layout is available on YouTube at \url{https://youtu.be/yDWdTUzTUMo}.}
A full description of the science capabilities of the array is found in
\citealt{Bowman:2013}.
The high imaging capability of the MWA enables us to calibrate and subtract foregrounds
based on sky models, leveraging the full capability of the observatory for the
EoR experiment. 
This strategy is in contrast to more targeted EoR experiments like PAPER, 
which utilize a highly redundant layout to enhance sensitivity. Redundant arrays can 
exploit symmetries of the instrument for quick calibration and analysis, but have poor 
point spread functions, making foregrounds more difficult to characterize.

@@

An example two dimensional residual power spectrum formed from the three hour golden 
data set is shown in Figure~\ref{fig:example_ps}. The bulk of the residual power is in the 
so-called foreground ``wedge", indicated with the diagonal black lines where the solid line 
corresponds to sky emission at the horizon, and the dashed lines corresponds to emission 
at the edge of the MWA field of view. Above the wedge we see horizontal lines of 
contamination due to the periodic frequency sampling function. These lines are often 
referred to as the coarse band harmonics. We see vertical streaks at high $k_{||}$ and high 
$k_{\perp}$. These are due to sparse $(u,v)$ sampling beyond the dense core of the MWA 
(starting $\sim70\lambda$), and the foreground power mixing to high spectral 
modes \citep{Bowman:2009}. Between the coarse band lines and to the left of the vertical streaks we see regions 
which contain both positive and negative (indicated by blue on the color bar) pixels. These 
regions are where we have successfully isolated the foregrounds and retained a noise-like 
EoR window in the three hour integration.

@@

The two-dimensional power spectrum is a useful figure of merit (FoM) as we improve and 
refine our analysis pipeline. Foregrounds and systematics often manifest with characteristic 
shapes in this space, enabling us to diagnose problems and quantify improvements 
\citep{Morales:2012}. We use the MWA 3 hour ``golden'' data set from August 23, 2013, to 
repeatedly form power spectra to test and refine our analysis. While an exhaustive catalog 
of these improvements is outside the scope of this paper, we 
demonstrate the utility of the 2D power spectrum as a FoM with a few key techniques we 
have employed in our analysis.

@@

J.~L.~B.~Line, et al (2016, in review) performed similar analysis, but with an emphasis on the
positional accuracy of catalog entries. By using simulations of visibilities they were able to 
isolate the effect of using perfect versus offset source positions. In their simulations source 
position offsets of $14\%$ of the synthesized
beam-width had significant impact on foreground subtraction, confirming the
importance of a complete and accurate point source catalog in order to retain a clean EoR
window.

@@

Ultimately we need model visibilities to subtract from our data. However, rather than storing 
a model for all observations (different pointings and phase centers), we find it most simple 
to treat the model as a single HEALPix image which can be used to create model visibilities 
for each independent observation. FHD treats each pixel in the image as a point source at 
the pixel location with total flux density equal to the surface brightness of the diffuse image times the area 
of the pixel and creates model visibilities for each snapshot in the same way that it imports 
a catalog of point sources. This is similar to the strategy employed by 
\citealt{Thyagarajan:2015} to model diffuse structure.

@@

\section{A Deep Integration}\label{sec:deep}
Up to this point we have only discussed testing and analysis on the three hour golden data 
set. We next turn towards a deeper integration, incorporating the techniques described above. 
We start with all MWA observations taken on the EoR0 field at 182~MHz between August 
23, 2013 and November 29, 2013. This includes 2,780 snapshots, or about 86.5 hours of 
data. We first make data quality cuts, then form power spectra.

@@

We use the jackknife method to filter out bad data and detect patterns. This involves dividing
the data into many subgroups and forming power spectra. The goal is to find observational
parameters which affect the quality of the data, such as the day or time of night.
A powerful grouping is to divide the data into the observation day 
and ``pointing" (which direction the antennas were pointed while tracking the EoR0 field). 
We show an example of one day's worth of per-pointing power spectra in Figure~
\ref{fig:jackknife}. The pointings are labeled sequentially with $-5$ corresponding to five 
pointing steps prior to zenith transit, $0$ corresponding to zenith, and $+4$ corresponding 
to four steps after zenith. 
This numbering scheme is shown graphically in Figure~\ref{fig:pointing}.
Early in the night ($-5$ through $-3$) the bulk of the Galactic plane 
was still above the horizon, and despite being very far from our primary field of view, highly 
contaminated our observations through the sidelobes of the instrument. We then have 
relatively well-behaved pointings, with the EoR window dominated by noise, until the final 
pointing of the night when we can see evidence of the Galaxy rising again indicated by 
strong lines of power at the edge of the foreground wedge. We saw the same contamination 
on all days of observation and decided to cut all $-5$, $-4$, $-3$, and $+4$ pointings from 
our data set. In principle it may be possible to model the Galactic plane well enough to 
account for its presence, but we leave this to future work.

@@

Motivated by the manual classification done with power spectrum jackknives, we developed 
another metric to quickly predict power spectrum quality for each snapshot in our data set. 
We do this by forming a delay spectrum \citep{Parsons:2012b} from the raw, uncalibrated 
visibilities. We then calculate an estimated total EoR window power by integrating the 
power above the horizon line and below the first coarse band line. While we would ideally 
use calibrated visibilities for this metric, we have found that the uncalibrated power is 
strongly correlated to the calibrated power, and the cuts we make below are independent of 
calibration. This is encouraging because future analysis could use this metric before 
processing the snapshots, saving valuable computing resources\footnote{Because
calibration involves generating model visibilities for each snapshot, approximately half
the computational cost of the entire imaging pipeline is required to obtain 
calibration solutions.}.

@@

The 32 hour integrated residual power spectra for our three sub-bands and both 
instrumental polarizations are shown in Figure~\ref{fig:2d_ps_subband}. A number of 
features can be seen in these spectra. In all cases the foreground wedge is prominent, with 
extra power at large scales (low $k_{\perp}$). This is an indication that our diffuse model needs to 
improve for deeper subtraction.

@@

The lowest region of the EoR window, above the horizon line and around $k_{||} \approx 
0.25$ h Mpc$^{-1}$, contains purely positive bins, indicating non-noise-like power. The 
cause for this seeming leakage is yet unknown, and limits our integration in this 
analysis. One suspected origin of this contamination was individual snapshots with high 
contamination that slipped through our cuts. However, dividing the data into random 
subsets and comparing results indicated that there is not a small number of offending 
observations, instead the leakage appears to exist at a low level in all the data.

@@

In all bands and polarizations, we are heavily signal dominated in our most sensitive region
(low $k$). This is not surprising based on the 2D spectra we examined earlier. The gaps in
the data are due to the excised coarse band harmonics. We can see the cable reflection line
between the first two gaps in all polarizations and sub-bands. Between the coarse band
harmonic lines, especially at larger $k$, are bins which approach the noise level and are
consistent with zero (notably the North-South polarization for the low and mid bands). The
bin just above the first coarse band harmonic is also consistent with zero in the low band.
These regions are encouraging because they have the potential to continue integrating down
with more data without further analysis improvements. Further suppression of the coarse band harmonic and cable reflection lines
will also improve the limits presented here.

@@

The second limited region in our 1D power spectra is between the coarse band harmonics 
where our measured signal approaches the noise level. Additional data will likely reduce the 
noise, and therefore our upper limit, in this regime. However, until the leakage at low $k$ is 
understood, progress at higher $k$ is susceptible to running into the same systematic
with longer integrations.

Finally, we place our best upper limits in context with the 21cm EoR field as a whole. A 
direct comparison between measurements from different instruments is difficult due to the 
varying methods, redshifts, and scales probed. However, if we approximate the power 
spectrum in units of mK$^2$ to be roughly flat over the scales probed by the MWA, PAPER, 
and GMRT, we can glean a view of the current state of the field. The results from this 
analysis are a modest improvement over the previous best MWA results \citep{Dillon:2015}, 
which is not surprising given that this integration contains about ten times more data, but 
is systematic limited. Our results are also consistent with the 
40 hour GMRT limit of $\Delta^2\le6.15\times10^4$~mK$^2$ at $k=0.5\, h$~Mpc$^{-1}$ and $z=8.6$ \citep{Paciga:2013}, though 
probe significantly different periods of the EoR. The frontrunner in terms of an upper limit is 
the PAPER experiment \citep{Ali:2015, Jacobs:2015, Parsons:2014}. The PAPER-64 limit of 
$\Delta^2\le502$~mK$^2$ at $z=8.4$ from a 135 day observing campaign is about one and a half orders of magnitude lower 
than our best limit in power units, though again probing a significantly different redshift. 
The MWA analysis is rapidly improving as we uncover and mitigate systematics to realize
the full sensitivity of the telescope.
The red lines in Figure~\ref{fig:1d_ps} indicate the noise level possible 
to reach with these data if systematics can be overcome.

@@

While the analysis presented here has not reached full potential, it represents the deepest power spectrum 
integration to date produced by an imaging pipeline. Imaging involves many difficulties (e.g. 
efficient gridding and mapmaking, foreground modeling), but if systematics can be 
overcome, it has the potential to compete with other more targeted experiments and 
analysis styles. 
More broadly, imaging analyses will be necessary to perform cross-correlation studies
with complimentary probes such as galaxy surveys or other intensity mapping experiments
 \citep[e.g.][]{DeBoer:2016, Vrbanec:2016, Beardsley:2015, Silva:2015, Dore:2014, Lidz:2009}, which will ultimately unlock the full potential of 21cm observations.
Here we have identified a crucial region of power spectrum space that is 
currently contaminated, with suggestions for improvement in future analysis. With improved 
calibration techniques, primary beam models, and understanding of foregrounds, the MWA 
and other imaging analyses will be able to quickly approach the most competitive upper 
limits in the field.
